---
title: "ZIPBN: Zero-inflated Poisson Bayesian Networks"
author: "Junsouk Choi"
date: "Dec 7, 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ZIPBN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In this vignette, we will give a tutorial on using `ZIPBN`. 
`ZIPBN` fits zero-inflated Poisson Bayesian networks (ZIPBN) for zero-inflated count data,  such as scRNA-seq data, using Markov chain Monte Carlo (MCMC).
The ZIPBN models deal with conditional dependencies of variables of zero-inflated count data. 
The model and its MCMC implementation are intermediate results of on-going work on scalable Bayesian networks for the scRNA-seq data.
Therefore, any manuscripts are not available yet, which explain the ZIPBN model and how to implement it. 
Before showing how to use `ZIPBN` to make a network analysis on zero-inflated count data, we briefly introduce the ZIPBN model and its MCMC schema.

```{r setup}
library(ZIPBN)
```




